Network based multifactorial modelling microRNA-target interaction.

### 2019-04-08 article.Rmd
check

## Methods

### miRNA:target ratio theoretical action in network based ceRNA model

We developed a novel network based approach that can use to simulation of gene regulation after any perturbation at miRNA:target interactions. If a perturbation or change induces on graph, whole interacted nodes vary because of changes in miRNA:target ratios. The system was assumed as steady state that miRNAs affect according to the ratio of the individual target to all targets (\autoref{fig1}a, expression values shown as integer for comprehensible sampling). When expression of gene 2 (G2) increased to two fold (\autoref{fig1}b), distribution of miRNA (M1) expression changes (\autoref{fig1}c). Consequently, expression of common target (G4) of miRNA1 and miRNA2 also varies. The change at expression of common target triggers to changes of proportional distribution of miRNA2 (\autoref{fig1}d).

![Shematic presentation of mechanism of network based model. a) In steady state, miRNAs (M: triangle grey) repress targets (G: circle coconut) according to proportion of their targets' expression. b-c-d) MiRNA expression distributions to targets vary when expression of one of the targets perturbed. ](Fig1.jpg){#fig1}

The model was developed by making some assumptions. Transcription rates and degradation rates of miRNAs were assumed as stationary and equal. Targets also have stabil transcription and degradation rates, at the same time they are equal. The factors that could be affect the miRNA:target interaction are disregarded in this approach. It is considered that miRNAs degradates or inhibits them via any mechanisms after binding to target while the miRNAs were used reversibly. In initial conditions, gene expression amounts in system have been considered after repression activity of miRNAs. If a change occurs in the system, variation on miRNA proportional distribution are reflected as change of repression activity.

\begin{equation} 
    Eff_{gi}= C_m * C_{gi}/\sum_{1}^{i} C_g \tag{1}\label{eq:1}
\end{equation}

\begin{equation}
   R_{gi}= Eff_{i1}+ Eff_{i2} ... \tag{2}\label{eq:2}
\end{equation}

@@

![Calculations for determine to miRNA binding and repression efficiency. $G$, Gene; $M$, miRNA; $STE$, seed type effect; $RE$, Region Effect; $E$, Energy; $STE’$, normalized values of seed type efficiency coefficient; $RE’$, normalized values of region efficiency coefficient; $E’$, normalized values of energy coefficient.](fig2.jpg){#fig2}

@@

Interactions between miRNAs and their targets can be affected from different parameters. For this reason, we improved model with factors that can affect to miRNA activity. In our approach, the parameters are classified as binding and efficiency factors and all of these are evaluated in their groups. Binding factors determine interaction between miRNA and target. The efficiency factors decide functionality of binding. In the literature, binding free energy [@cao_predicting_2012 ; @helwak_mapping_2013] and seed type [@werfel_preferential_2017]in miRNA:target interactions are represented as binding affinity. Efficiency factors determine how many of microRNA:target complexes will result in inhibition. For example, binding region on the target shows whether the target can be inhibited (functionality)[@hausser_analysis_2013; @helwak_mapping_2013]. Both of binding and efficiency factors are evaluated in the groups and normalized according to equation \eqref{eq:3}; each factor is normalized based on maximum variables that found in their groups.

\begin{equation}
F`= F/F_{max} \tag{3}\label{eq:3}
\end{equation}

The normalized values of factors take into account to determine binding activity and miRNA efficiency on targets (\autoref{fig2}).  Binding affinities (activity, $Eff$) of miRNAs on each individual gene are calculated as shown in equation \eqref{eq:4}; $C_m$, miRNA expression in the group; $C_g$, Gene expression; $C_{gi}$, individual gene expression; gi, individual gene; g, whole genes in a group (\autoref{fig2}c).

\begin{equation}
Eff_{gi}= C_m * E`_{gi}*STE`_{gi}*C_{gi}/(\sum_{0}^{i} E`_{gi}*STE`_{gi}*C_{gi}) \tag{4}\label{eq:4}
\end{equation}

\begin{equation}
Eff_{gi}= Eff_{gi}*RE`_{gi} \tag{5}\label{eq:5}
\end{equation}

@@

### Defined functions for ceRNA models

We defined the functions that can be used with R programming. Briefly, these functions process a given miRNA:gene dataset and convert to graph object. All values that are significant in miRNA:target interactions are stored in edge variables and processed with formulations that are given in previous sections. The functions and steps of approach are explained as following \autoref{fig3}:
1-2. Convertion of dataset: priming_graph function process the given dataset that includes competing elements in first variable and repressive element in second variable. If the affinity and/or degradation factors are specified in the function, factors are taken into account, are processed with defaults in vice versa. The formulations that are given in equations (i-v) are performed in this function. This step gives the graph object which contains efficiency values of miRNA:competing target pairs in steady-state in terms of amount. It is assumed that the initial target amounts in the dataset is observed after the reppressive activity of miRNAs in steady-state. 
2-3. Transition of variables in graph: In the previous step, the calculations are performed in the edge variables of the graph object. However, the graph object allows to use node variables, while the node features are handled to the graph. In this direction, update_nodes function carries the amount values to node variables This step must be applied with “once” option because it is primary process.
3-4. Trigger change in graph: The dataset are assumed as steady-state in previous step and the efficieny coefficients are calculated according to this acceptance. In the network that is found in steady-state conditions, the change is applied to the graph object for disturbtion of steady-state. To provide the disturbtion in the network the workflow offer two methods: update_variables and update_how. The fist, a new dataset that is contained competing and repressive element names and current values of these can be processed with update_variables. The second option, the amount of the given node name in update_how function can be changed according to “how” argument.
4-5. Updating current values of variables: After variables updating in edge varibles, these are carried to node variables. Current and previous values of variables are stored as node variables with update_variables function.
5-6. Simulation of competing behavior of targets: After the change in the steady- state conditions, the network elements try to gain steady-state again. This process progresses as repeating of regulations after the spreading the changes in the network. In this step, simulation of regulations according to given cycle count in simulate function is applied. After each simulation cycle, the miRNA repression values are re-calculated and the current values of competing elements are found and saved. The process is performed in the edge data and at the same time outputs of the calculations are carried from edge to node data.
The node elements in the dataset are handled as two type; repressives (miRNAs) and competings (targets). It is assumed in approach that while targets are degrading or inhibiting by miRNAs continuously, miRNAs reversibly used. If the trigger of the network is a miRNA, it maintains the current value of amount that provides by user. On the contrary, it tries to help this process to provide steady-state through the regulations on its amount, if a competing element is used as a trigger. The functions that are used in the approach are developed with R programming so as can be used with other packages. These are can be found in the github repository [ceRNAnetsim github page](https://github.com/selcenari/ceRNAnetsim) and improved with contributions of others.

@@

We developed a model approach and workflow for competitive ceRNA regulation in this study. The basic mode of miRNA repression activity has been based on miRNA and target abundance in various researches[@arvey_target_2010; @denzler_assessing_2014]. For this reason, we performed first application as considering only abundance of miRNAs and targets (\autoref{fig1}). After the increasing in a gene expression value, expression values of other genes also changed slightly in comparison with alteration in perturbation started gene node. Different studies have shown that the other gene expression values also rise differently, if a gene abundance increases in ceRNA system [@lai_understanding_2016; @salmena_cerna_2011; @tay_multilayered_2014]. It was observed that primary neighborhoods of the trigger gene change with the same ratio, but cooperative efficiency of miRNAs on common target causes differently change in common target (\autoref{fig1}d). The common target in system displayed like a trigger for the other group and induced changes of expression values of genes on the other group. Therefore, effect of primary change of the gene (G2) on genes of the other group was observed slightly (\autoref{fig1}d). In addition, like shown by ceRNA hypothesis model of Ala et al., after the increasing of a gene expression (G2), the miRNA that is found in the same group (M1) tended to be more repressive on this target. This caused to decrease in increased expression value of primer triggering gene.
When the expression of a gene (G2) was increased to two fold, we observed that the expression changes were more evident. So, we considered that high miRNA:target ratio results in stronger miRNA repression activity. This results also are coherent with previous reports that were offered by Arvey et al. and Denzler et al.[@arvey_target_2010; @bosson_endogenous_2014; @denzler_assessing_2014]. When the simulation was ran in sample dataset that includes lower target abundance, the first response of primary neihgborhood of the same group was determined as similar with in system that have high target abudance. But regulations of ceRNAs were observed differences in the folowing steps of simulation. In overall state in the end of simulation, we established that all regulations more prominent when compared with low miRNA:target ratio datset. The model appeared the importance of mRNA:target ratio in regulation of competing RNAs.

@@

It has been reported in several studies that miRNA regulatory interactions also is affected different parameters. For example, Xu et. al have investigated the importance of seed pairing type between miRNAs and their targets and target site location by using proteomic dataset [@xu_characterization_2014]. They have proposed that the features of binding between miRNA and target can be critical efficiency of miRNA and can be used to find effective miRNA molecules. In addition, binding energy between miRNA and targets also could be determinant for miRNA efficiency. It has been reported by Breda et. al. that strength of miRNA:target interactions is depended on binding energy of complexes [@breda_quantifying_2015]. As similar to these, Bosson et al. have associated affinity and seed pairing of miRNA target pairs and suggested that number of canonical base pairing is correlated with affinity. For these reasons, we improved the model with optional interaction parameters that could be useful for understanding of miRNA repression activity. We explained the model in the same sample dataset with using different factors and simulated via the same change (\autoref{fig4}). When the factors were taken into account in the system, miRNA efficiencies varied as shown in \autoref{fig4}a. Although the miRNA target ratios in initial states were same in comparison with the sample dataset without factors, efficiency of binding and repression have changed. When expression of a gene (G2) increased, expression values of all genes also change differentially because of contribution of efficiency factors (\autoref{fig4} b,c,d).
In regulation of miRNA target sample system (\autoref{fig4}), miRNA (M1) repressive efficiency on the primary triggering gene (G2) is low in steady-state. So it has been observed that the regulatory activities of miRNA efficiencies on the targets are weak after the increase of miRNA target G2.  When we ran to model with two fold increasing expression value of common target (G4), the changes of other gene expressions were observed more prominent. Furthermore, folding of expression of target gene that has strong miRNA repression efficiency also resulted in evident perturbation.
The defined miRNA’ number is very low in comparison with defined protein coding genes’ number. So, this results in that the miRNA amounts in the system must be high or targeted RNAs’ amount must be low for effective miRNA activity. For these reasons, we tried to simulate a real dataset which contains thousands of genes and hundreds of miRNAs. Briefly, we have obtained the datasets and combined these via R programming. We have applied the our approach into integrated dataset using a selected gene. Firstly we observed the activities in the iterations and determined behaviours of the network after simulation. As expected, we have observed that the the huge dataset, compared to those containing a few genes and miRNAs, exhibited cumbersome behaviours and complexity to gain steady-state. We realized that a small change in a system with a large change actually brought many arrangements. This can be explained with the competing groups. Because the genes are targeted with many common miRNAs, and this results in that the common targets exhibits competing behaviours in different groups. On the other hand, the changes of gene expressions were very different. While the change in some of genes were highly evident, somes have changed slightly. So, we have compared two datasets: the tumor gene expression and simulation results. When we analysed comparison of two datasets, we observed that the results of simulation is not convenient with tumor tissue expression values. The simulation results were not expected to be concordant with tumor tissue expression values because of a single gene could not be responsible for gene regulation in all tissue. Besides, the all other factors such as up/down regulation of miRNAs or other genes were ignored. Of course, it would be more useful to test a dataset that includes gene expression values before and after the regulation of a gene. When we evaluated the suitability of approach, the approach allows different miRNA targets to be considered together and can work with different regulatory elements due to the network structure. 

>>>>

## Methods
### miRNA:target ratio theoretical action in network based ceRNA model
We developed a novel network based approach that can use to simulation of gene regulation after any perturbation at miRNA:target interactions. If a perturbation or change induces on graph, whole interacted nodes vary because of changes in miRNA:target ratios. The system was assumed as steady state that miRNAs affect according to the ratio of the individual target to all targets (\autoref{fig1}a, expression values shown as integer for comprehensible sampling). When expression of gene 2 (G2) increased to two fold (\autoref{fig1}b), distribution of miRNA (M1) expression changes (\autoref{fig1}c). Consequently, expression of common target (G4) of miRNA1 and miRNA2 also varies. The change at expression of common target triggers to changes of proportional distribution of miRNA2 (\autoref{fig1}d).

![Shematic presentation of mechanism of network based model. a) In steady state, miRNAs (M: triangle grey) repress targets (G: circle coconut) according to proportion of their targets' expression. b-c-d) MiRNA expression distributions to targets vary when expression of one of the targets perturbed.](Fig1.jpg){#fig1}

The model was developed by making some assumptions. Transcription rates and degradation rates of miRNAs were assumed as stationary and equal. Targets also have stabil transcription and degradation rates, at the same time they are equal. The factors that could be affect the miRNA:target interaction are disregarded in this approach. It is considered that miRNAs degradates or inhibits them via any mechanisms after binding to target while the miRNAs were used reversibly. In initial conditions, gene expression amounts in system have been considered after repression activity of miRNAs. If a change occurs in the system, variation on miRNA proportional distribution are reflected as change of repression activity.
\begin{equation} 
    Eff_{gi}= C_m * C_{gi}/\sum_{1}^{i} C_g \tag{1}\label{eq:1}
\end{equation}
\begin{equation}
   R_{gi}= Eff_{i1}+ Eff_{i2} ... \tag{2}\label{eq:2}
\end{equation}

@@

![Calculations to determine of miRNA binding and repression efficiency. $G$, Gene; $M$, miRNA; $STE$, seed type effect; $RE$, Region Effect; $E$, Energy; $STE’$, normalized values of seed type efficiency coefficient; $RE’$, normalized values of region efficiency coefficient; $E’$, normalized values of energy coefficient.](fig2.jpg){#fig2}

@@

Interactions between miRNAs and their targets can be affected from different parameters. For this reason, we improved model with factors that can affect to miRNA activity. In our approach, the parameters are classified as binding and efficiency factors and all of these are evaluated in their groups. Binding factors determine interaction between miRNA and target. The efficiency factors decide functionality of binding. In the literature, binding free energy [@cao_predicting_2012 ; @helwak_mapping_2013] and seed type [@werfel_preferential_2017] in miRNA:target interactions are represented as binding affinity. Efficiency factors determine how many of microRNA:target complexes will result in inhibition. For example, binding region on the target shows whether the target can be inhibited (functionality)[@hausser_analysis_2013; @helwak_mapping_2013]. Both of binding and efficiency factors are evaluated in the groups and normalized according to equation \eqref{eq:3}; each factor is normalized based on maximum variables that found in their groups.
\begin{equation}
F`= F/F_{max} \tag{3}\label{eq:3}
\end{equation}

The normalized values of factors take into account to determine binding activity and miRNA efficiency on targets (\autoref{fig2}). Binding affinities (activity, $Eff$) of miRNAs on each individual gene are calculated as shown in equation \eqref{eq:4}; $C_m$, miRNA expression in the group; $C_g$, Gene expression; $C_{gi}$, individual gene expression; gi, individual gene; g, whole genes in a group (\autoref{fig2}c). 
\begin{equation}
Eff_{gi}= C_m * E`_{gi}*STE`_{gi}*C_{gi}/(\sum_{1}^{i} E`_{gi}*STE`_{gi}*C_{gi}) \tag{4}\label{eq:4}
\end{equation}
\begin{equation}
Eff_{gi}= Eff_{gi}*RE`_{gi} \tag{5}\label{eq:5}
\end{equation}

@@

### Defined functions for ceRNA models
We defined the functions that can be used with R programming. Briefly, these functions process a given miRNA:gene dataset and convert to graph object. All values that are significant in miRNA:target interactions are stored in edge variables and processed with formulations that are given in previous section. The functions and steps of approach are explained as following (\autoref{fig3}) :
*Convertion of dataset*: priming_graph function processes the given dataset that includes competing elements in first variable and repressive element in second variable. If the affinity and/or degradation factors are specified in the function, factors are taken into account, are processed with defaults in vice versa. The formulations that are given in equations \eqref{eq:1}-\eqref{eq:5} are performed in this function. This step gives the graph object which contains efficiency values of miRNA:competing target pairs in steady-state in terms of amount. It is assumed that the initial target amounts in the dataset is observed after the reppressive activity of miRNAs in steady-state. *Transition of variables in graph*: In the previous step, the calculations are performed in the edge variables of the graph object. However, the graph object allows to use node variables, while the node features are handled to the graph. In this direction, update_nodes function carries the amount values to node variables This step must be applied with “once” option because it is primary process.
*Trigger change in graph*: The dataset are assumed as steady-state in previous step and the efficieny coefficients are calculated according to this acceptance. In the network that is found in steady-state conditions, the change is applied to the graph object for disturbtion of steady-state. To provide the disturbtion in the network the workflow offer two methods: update_variables and update_how. The first, a new dataset that is contained competing and repressive element names and current values of these can be processed with update_variables. The second option, the amount of the given node name in update_how function can be changed according to “how” argument.
*Updating current values of variables*: After variables updating in edge varibles, these are carried to node variables. Current and previous values of variables are stored as node variables with update_variables function.
*Simulation of competing behavior of targets*: After the change in the steady-state conditions, the network elements try to gain steady-state again. This process progresses as repeating of regulations after the spreading the changes in the network. In this step, simulation of regulations according to given cycle count in simulate function is applied. After each simulation cycle, the miRNA repression values are re-calculated and the current values of competing elements are found and saved. The process is performed in the edge data and at the same time outputs of the calculations are carried from edge to node data.
The node elements in the dataset are handled as two type; repressives (miRNAs) and competings (targets). It is assumed in approach that while targets are degrading or inhibiting by miRNAs continuously, miRNAs reversibly used. If the trigger of the network is a miRNA, it maintains the current value of amount that provides by user. On the contrary, it tries to help this process to provide steady-state through the regulations on its amount, if a competing element is used as a trigger. The functions that are used in the approach are developed with R programming so as can be used with other packages. These are can be found in the github repository [ceRNAnetsim github page](https://github.com/selcenari/ceRNAnetsim) and improved with contributions of others.

@@

We have developed a model approach and workflow for competitive ceRNA regulation in this study. The basic mode of miRNA repression activity has been based on miRNA and target abundance in various researches[@arvey_target_2010; @denzler_assessing_2014]. For this reason, we performed first application as considering only abundance of miRNAs and targets (\autoref{fig1}). After the increasing in a gene expression value, expression values of other genes also changed slightly in comparison with alteration in perturbation started gene node. Different studies have shown that the other gene expression values also rise differently, if a gene abundance increases in ceRNA system [@lai_understanding_2016; @salmena_cerna_2011; @tay_multilayered_2014]. It was observed that primary neighborhoods of the trigger gene change with the same ratio, but cooperative efficiency of miRNAs on common target causes differently change in common target (\autoref{fig1}d). The common target in system has displayed like a trigger for the other group and induced changes of expression values of genes on the other group. Therefore, effect of primary change of the gene (G2) on genes of the other group was observed slightly (\autoref{fig1}d). In addition, like shown by ceRNA hypothesis model of Ala et al., after the increasing of a gene expression (G2), the miRNA that is found in the same group (M1) tended to be more repressive on this target. This has caused to decrease in increased expression value of primer triggering gene. When the expression of a gene (G2) was increased to two fold, we observed that the expression changes were more evident. So, we have considered that high miRNA:target ratio results in stronger miRNA repression activity. This results also are coherent with previous reports that were offered by Arvey et al. and Denzler et al.[@arvey_target_2010; @bosson_endogenous_2014; @denzler_assessing_2014]. When the simulation was run in sample dataset that includes lower target abundance, the first response of primary neihgborhood of the same group was determined as similar with in system that have high target abudance. But regulations of ceRNAs were observed differences in the folowing steps of simulation. In overall state in the end of simulation, we have established that all regulations more prominent when compared with low miRNA:target ratio dataset. The model has appeared the importance of mRNA:target ratio in regulation of competing RNAs.

@@

It has been reported in several studies that miRNA regulatory interactions also is affected different parameters. For example, Xu et. al have investigated the importance of seed pairing type between miRNAs and their targets and target site location by using proteomic dataset [@xu_characterization_2014]. They have proposed that the features of binding between miRNA and target can be critical efficiency of miRNA and can be used to find effective miRNA molecules. In addition, binding energy between miRNA and targets also could be determinant for miRNA efficiency. It has been reported by Breda et. al. that strength of miRNA:target interactions is depended on binding energy of complexes [@breda_quantifying_2015]. As similar to these, Bosson et al. have associated affinity and seed pairing of miRNA target pairs and suggested that number of canonical base pairing is correlated with affinity. For these reasons, we have improved the model with optional interaction parameters that could be useful for understanding of miRNA repression activity. We have explained the model in the same sample dataset with using different factors and simulated via the same change (\autoref{fig4}). When the factors were taken into account in the system, miRNA efficiencies varied as shown in \autoref{fig4}a. Although the miRNA target ratios in initial states were same in comparison with the sample dataset without factors, efficiency of binding and repression have changed. When expression of a gene (G2) increased, expression values of all genes also change differentially because of contribution of efficiency factors (\autoref{fig4} b,c,d).

In regulation of miRNA target sample system (\autoref{fig4}), miRNA (M1) repressive efficiency on the primary triggering gene (G2) is low in steady-state. So it has been observed that the regulatory activities of miRNA efficiencies on the targets are weak after the increase of miRNA target G2. When we ran to model with two fold increasing expression value of common target (G4), the changes of other gene expressions have observed more prominent. Furthermore, folding of expression of target gene that has strong miRNA repression efficiency also resulted in evident perturbation.

The defined miRNA’ number is very low in comparison with defined protein coding genes’ number. So, this results in that the miRNA amounts in the system must be high or targeted RNAs’ amount must be low for effective miRNA activity. We have tried to simulate a real dataset which contains thousands of genes and hundreds of miRNAs. Briefly, we have obtained the datasets and combined these via R programming. We have applied the our approach into integrated dataset using a selected gene. Firstly, we have observed the activities in the iterations and determined behaviours of the network after simulation. As expected, we have observed that the big dataset, compared to those containing a few genes and miRNAs, exhibited cumbersome behaviours and complexity to gain steady-state. We realized that a small change in a system with a large change actually brought many arrangements. This can be explained with the competing groups. Because the genes are targeted with many common miRNAs, and this results in that the common targets exhibits competing behaviours in different groups. On the other hand, the changes of gene expressions were very different. While the change in some of genes were highly evident, somes have changed slightly. So, we have compared two datasets: the tumor gene expression and simulation results. When we analysed comparison of two datasets, we have observed that the results of simulation is not convenient with tumor tissue expression values. The simulation results were not expected to be concordant with tumor tissue expression values because of a single gene could not be responsible for gene regulation in all tissue. Besides, the all other factors such as up/down regulation of miRNAs or other genes were ignored. Of course, it would be more useful to test a dataset that includes gene expression values before and after the regulation of a gene. When we evaluated the suitability of approach, the approach allows different miRNA targets to be considered together and can work with different regulatory elements due to the network structure.
